Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-11-13 mnras.tex
Added MNRAS version

\documentclass[apj,onecolumn]{emulateapj}
%\documentclass[12pt,preprint]{aastex}

\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{natbib}
\usepackage[section] {placeins}
\bibliographystyle{apj}
\citestyle{apj}

@@

\shorttitle{AMR SIMULATIONS WITH ADAPTIVE RAY TRACING}
\shortauthors{WISE ET AL.}

\title{Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement
  Simulations with Adaptive Ray Tracing}

\author{John H. Wise\altaffilmark{1}} 
\affil{Department of Astrophysical Sciences, Princeton University,
  Peyton Hall, Ivy Lane, Princeton, NJ 08544}

\author{Ji-hoon Kim and Tom Abel} 
\affil{Kavli Institute for Particle Astrophysics and Cosmology,
  Stanford University, Menlo Park, CA 94025}

\altaffiltext{1}{Hubble Fellow}
\email{jwise@astro.princeton.edu}

@@

  We describe a photon-conserving numerical method that solves the
  radiative transfer equation, using a spatially-adaptive ray tracing
  scheme, and its parallel implementation into the adaptive mesh
  refinement (AMR) code, Enzo.  By coupling the solver with the energy
  equation and non-equilibrium chemistry network, our radiation
  hydrodynamics framework can be utilized to study a broad range of
  astrophysical problems, such as stellar and black hole (BH)
  feedback.  Inaccuracies can arise from large timesteps and poor
  sampling, therefore we devised an adaptive time-stepping scheme and
  a fast approximation of the optically-thin radiation field with
  multiple sources.  We test the method with several radiative
  transfer and radiation hydrodynamics tests that are given in
  \citet{RT06, Iliev09}.  We further test our method with more
  dynamical situations, for example, the propagation of an ionization
  front through a Rayleigh-Taylor instability, time-varying
  luminosities, and collimated radiation.  The test suite also
  includes an expanding \ion{H}{2} region in a magnetized medium,
  utilizing the newly implemented magnetohydrodynamics module in
  \enzo.  This method linearly scales with the number of point sources
  and number of grid cells.  To combat this, we use a novel method of
  merging rays at large radius, using a tree method, which we briefly
  describe.  Our implementation is scalable to 512 processors on
  distributed memory machines and can include radiation pressure and
  secondary ionizations from X-ray radiation.  It is included in the
  newest public release of Enzo.

@@

\keywords{cosmology --- methods: numerical --- hydrodynamics ---
  radiative transfer}

@@

Radiative transfer is ubiquitous in many astrophysical problems,
encompassing topics such as stellar atmospheres, the interstellar
medium (ISM), star formation, galaxy formation, supernovae (SNe) and
cosmology.  Thus it is a well-studied problem
\citep[e.g.][]{Mathews65, Rybicki, Mihalas84, Yorke86}; however, its
treatment in multi-dimensional calculations is difficult because of
the dependence on seven variables --- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  In the general case, one must consider all radiation
sources at every point in calculation.

Fortunately problems can be simplified by exploiting symmetries and
special cases.  The first case was considered by \citet{Stroemgren39}
for a radiation source ionizing a static uniform neutral medium, where
recombinations balance photo-ionizations.  He derived a radius of an
\ion{H}{2} region,
%
\begin{equation}
  \label{eqn:Rstr}
  R_s = \left(\frac{3\dot{N}_\gamma}{4\pi \alpha_{\rm B} n_H^2}\right)^{1/3},
\end{equation}
where $\dot{N}_\gamma$ is the ionizing photon luminosity, $\alpha_{\rm
  B}$ is the recombination rate, and $n_H$ is the ambient hydrogen
number density.  Furthermore he found that the delineation between the
neutral and ionized medium to be approximately the mean free path of
the ionizing radiation.  His seminal work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionizing radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionization front,
then this over-pressurized region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics.  A
decade later, the first radiation hydrodynamical numerical models of
\ion{H}{2} regions in spherical symmetry and plane-parallel ionization
fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionization front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  In R-type
fronts, the ionized gas density is higher than the neutral gas
density, and in D-type fronts, the opposite is true.  R-type fronts
travel supersonically with respect to the neutral gas, whereas D-type
fronts are subsonic.  Furthermore ``weak'' and ``strong'' R-type fronts
move supersonically and subsonically with respect to the ionized gas,
respectively.  The same terminology conversely applies to D-type
fronts.  ``Critical'' fronts are defined as moving exactly at the
sound speed.  These works established the evolutionary track of an
expanding \ion{H}{2} illuminated by a massive star in a uniform
medium:
%
\begin{enumerate}
\item Weak R-type---When the star (gradually) starts to shine, the
  ionization front will move supersonically through the ambient
  medium.  The gas is heated and ionized, but otherwise left
  undisturbed.  This stage continues until $r \sim 0.02R_s$.
\item Critical R-type---As the ionization front moves outwards, it
  begins to slow because of the geometric dilution of the radiation.
  It becomes a critical R-type front, which is equivalent to an
  isothermal shock in the neutral gas.
\item Strong and weak D-type---The front continues to slow, becoming a
  strong D-type front, and then a critical D-type front.  From this
  point forward, it is moving subsonically with respect to the ionized
  gas, i.e. a weak D-type front.  Thus sound waves can travel across
  the ionization front and form a shock.  The ionization front
  detaches from the shock, putting the shock ahead of the ionization
  front.
\item Expansion phase---After the shock forms, the \ion{H}{2} region
  starts to expand, lowering the interior density and thus the
  recombination rates.  This increases the number of photons available
  for ionizing the gas.  The sphere expands until it reaches pressure
  equilibrium with the ambient medium at $r \sim 5R_s$.
\end{enumerate}
%

In the 1970's and 1980's, algorithmic and computational advances
allowed numerical models to be expanded to two dimensions, mainly
using axi-symmetric to simplify the problem
\citep[e.g.][]{Bodenheimer79, Sandford82, Yorke83}.  One topic that was
studied extensively were champagne flows.  Here the source is embedded
in an overdense region, and the \ion{H}{2} region escapes from this
region in one direction.  The interface between the ambient and dense
medium was usually set up to be a constant pressure boundary.  When
the ionization front passes this boundary, the dense, ionized gas is
orders of magnitude out of pressure equilibrium as the temperatures on
both sides of initial boundary are within a factor of a few.  In
response, the gas is accelerated outwards in this direction, creating
a fan-shaped outflow.

Only in the past 15 years, computational resources have become large
enough, along with further algorithmic advances, to cope with the
requirements of three-dimensional calculations.  There are two popular
methods to solve the radiative transfer equation in three-dimensions:
%
\begin{itemize}
\item Moment methods---The angular moments of the radiation field can
  describe its angular structure, which are related to the energy
  energy, flux, and radiation pressure \citep{Auer70, Norman98}.
  These have been implemented in conjunction with short
  characteristics \citep[2D]{Stone92_RHD}, with long characteristics
  \citep{Finlator09}, with a variable Eddington tensor in the
  optically thin limit \citep[OTVET;][]{Gnedin01_OTVET, Petkova09},
  and with an M1 closure relation \citep{Gonzalez07, Aubert08}.
  Moment methods have the advantage of being fast, being independent
  of the number of radiation point sources.  However, they are
  diffusive and result in incorrect shadows in some situations.
\item Ray tracing---Radiation can be propagated along rays that extend
  through the computational grid \citep[e.g.][]{Razoumov99, Abel99_RT,
    Ciardi01, Sokasian01, Whalen06, Rijkhorst06, Mellema06, Alvarez06,
    Trac07, Krumholz07_ART, Paardekooper10} or particle set
  \citep[e.g.][]{Susa06, Johnson07, Pawlik08, Pawlik10, Altay08,
    Hasegawa09}.  In general, these methods are very accurate but
  computationally expensive because the radiation field must be well
  sampled by the rays with respect to the spatial resolution of the
  grid or particles.
\end{itemize}
%
Until the mid-2000's the vast majority of the three-dimensional
calculations were performed with static density distributions.  One
example is calculating cosmological reionization by post-processing of
density fields from N-body simulations \citep{Ciardi01, Sokasian01,
  McQuinn07, Iliev06, Iliev07}.  Any hydrodynamical response to the
radiation field was thus ignored.  Several radiative transfer codes
were compared in four purely radiative transfer tests in
\citet[hereafter RT06]{RT06}.  Only recently has the radiative
transfer equation been coupled to the hydrodynamics in
three-dimensions.  In the second comparison paper \citep[hereafter
  RT09]{Iliev09}, results from these radiation hydrodynamics codes
were compared.  Even more rare are ones that couple it with
magneto-hydrodynamics \citep[e.g.][]{Krumholz07_ART}.  The tests in
RT06 and RT09 were kept relatively simple to ease the comparison.

In this paper, we present our implementation, Enzo+Moray, of adaptive
ray tracing \citep{Abel02_RT} in the cosmological hydrodynamics
adaptive mesh refinement (AMR) code, \enzo~\citep{BryanNorman1997,
  OShea2004}.  The radiation field is coupled to the hydrodynamics
solver at small timescales, enabling it to study radiation
hydrodynamical problems.  We have used this code to investigate the
growth of an \ion{H}{2} region from a 100\Ms~Population III (Pop III)
star \citep{Abel07}, the early stages of reionization from Pop III
stars \citep{Wise08_Reion}, radiative feedback on the formation of
high redshift dwarf galaxies \citep{Wise08_Gal}, ultraviolet radiation
escape fractions from dwarf galaxies before reionization
\citep{Wise09}, negative radiative feedback from accreting Pop III
seed black holes \citep{Alvarez09}, and radiative feedback in
accreting supermassive black holes \citep[][in prep.]{Kim11}.  We have
included Enzo+Moray in the latest public release of
\enzo\footnote{http://lca.ucsd.edu/enzo}, and it is also coupled with
the newly added MHD solver in \enzo~\citep{Wang09}.

@@

\input table1.tex

@@

\begin{figure}[t]
  \plottwo{fig/covering.eps}{fig/calc_fc.eps}
  \caption{\label{fig:covering} (a) A two-dimensional illustration of
    the overlap between the beam associated with a ray $\gamma$ and a
    computational cell.  The ray has a segment length of $dr$ passing
    through the cell.  The covering area is denoted by dark grey,
    where the full area $(dr \times L_{\rm pix})$ is colored by the
    dark and light grey.  The photo-ionization and photo-heating rates
    should be corrected by this overlap fraction $f_c$.  (b)
    Annotation of quantities used in this geometric correction.}
\end{figure}

@@

We must be careful not the overestimate the radiation when multiple
rays enter a single cell.  In the case of a single radiation source,
the solution is simple --- only assign the cell the photo-ionization
and photo-heating rates when $k_{\rm ph} = 0$.  However in the case
with multiple sources, this is no longer valid, and we must sum the
flux from all optically thin sources.  Only one ray per source must
contribute to a single cell in this framework.  We create a flagging
field that marks whether a cell has already been touched by an
optically thin photon from a particular radiation source.  Naively, we
would be restricted to tracing rays from a single source at a time if
we use a boolean flagging field.  However we can trace rays for 32
sources at a time by using bitwise operations on a 32-bit integer
field.  For example in \texttt{C}, we would check if an optically thin
ray from the \texttt{n}-th source has propagated through cell
\texttt{i} by evaluating \texttt{(MarkerField[i] $\gg$ n \& 1)}.  If
false, then we can add the optically thin approximation
(Eq. \ref{eqn:optthin}) to the cell and set \texttt{MarkerField[i] |=
  (1 $\gg$ n);} to mark the cell.

@@

Molecular hydrogen can absorb photons in the Lyman-Werner bands
through the two-step Solomon process that are composed of 76
absorption lines ranging from 11.1 to 13.6 eV \citep{Stecher67,
  Dalgarno70}.  Each of these spectral lines can be modelled with a
typical exponential attenuation equation \citep{Ricotti01}, but
\citet{Draine96} showed that this self-shielding is well modeled with
the following relation to total \hh~column density
%
\begin{equation}
  \label{eqn:lwband}
  f_{\rm shield}(N_{\rm H2}) = \left\{ \begin{array}{l@{\quad}l}
      1 & (N_{\rm H2} \le 10^{14} \; \mathrm{cm}^{-2})\\
      (N_{\rm H2}/10^{14} \; \mathrm{cm}^{-2})^{-0.75} & (N_{\rm H2} >
        10^{14} \; \mathrm{cm}^{-2})
    \end{array} \right. .
\end{equation}
To incorporate this shielding function into the ray tracer, we store
the total \hh~column density and calculate the \hh~dissociation rate
by summing the contribution of all rays,
%
\begin{equation}
  \label{eqn:lwRT}
  k_{\rm diss} = \sum_{\rm rays} \frac{P \; \sigma_{\rm LW} \;
    \Omega_{\rm ray} \; r^2 \; dr}{A_{\rm cell} \; dV \; dt_P},
\end{equation}
where $\sigma_{\rm LW} = 3.71 \times 10^{18} \mathrm{cm}^2$ is the
effective cross-section of \hh~\citep{Abel97}.  To account for
absorption, we attenuate the photon number flux by
%
\begin{equation}
  \label{eqn:LWdP}
  dP = P [f_{\rm shield}(N_{\rm H2} + dN_{\rm H2}) - f_{\rm shield}(N_{\rm H2})],
\end{equation}
where $dN_{\rm H2}$ is the \hh~column density in the current cell.

@@

At the other end of the spectrum, a high-energy ($E_{\rm ph} \gsim
100$ eV) photon can ionize multiple neutral hydrogen and helium atoms,
and this should be considered in such radiation fields.
\citet{Shull85} studied this effect with Monte Carlo calculations over
varying electron fractions and photon energies up to 3 keV.  They find
that the excitation of hydrogen and helium and the ionization of
\ion{He}{2} is negligible.  The number of secondary ionizations of H
and He is reduced from the ratio of the photon and ionization energies
($E_{\rm ph} / E_i$) by a factor of
%
\begin{equation}
  Y_{\rm k,H} = 0.3908 (1 - x^{0.4092})^{1.7592},
\end{equation}
\begin{equation}
  Y_{\rm k,He} = 0.0554 (1 - x^{0.4614})^{1.6660},
\end{equation}
where $x$ is the electron fraction.  The remainder of the photon
energy is deposited into thermal energy that is approximated by
%
\begin{equation}
  Y_{\rm \Gamma} = 0.9971 [ 1 - (1 - x^{0.2663})^{1.3163} ]
\end{equation}
and approaches one as $x \rightarrow 1$.  Thus in gas with low
electron fractions, most of the energy results in ionizations of
hydrogen and helium, and in nearly ionized gas, the energy goes into
photo-heating.

@@

\begin{figure}[t]
  \plotone{fig/RTcoupling.eps}
  \caption{\label{fig:MainAlgorithm} Flow chart for the overall
    algorithm of the radiative transfer module in \enzo~that
    illustrates (1) the creation of photon packages, (2) ray tracing,
    (3) the transport of photon packages between AMR grids, and (4)
    coupling with the hydrodynamics.  The ray tracing algorithm, which
    is contained in the ``Trace Rays'' is detailed in Figure
    \ref{fig:RTalgorithm}.}
\end{figure}

\begin{figure}[t]
  \epsscale{1.1}
  \plotone{fig/RTalgorithm.eps}
  \caption{\label{fig:RTalgorithm} Flow chart for the ray tracing
    algorithm for one photon passing through a grid.  Note that only
    one step is needed in the routine to adaptively split rays.  The
    remainder is a typical ray tracing method.}
  \epsscale{1}
\end{figure}

@@

\step{3g} Add the effects of any optional physics modules
(\S\ref{sec:addphysics})---secondary ionizations from X-rays, Compton
heating from scattering, and radiation pressure.

@@

In our implementation, photon packages are mono-chromatic, i.e. energy
groups \citep[][Ch. 6]{Mihalas84}, and are assigned a photon type that
corresponds whether it is a photon that (1) ionizes hydrogen, (2)
singly ionizes helium, (3) doubly ionizes helium, (4) has an X-ray
energy, or (5) dissociates molecular hydrogen (Lyman-Werner
radiation).  One disadvantage of mono-chromatic rays is the number of
rays increase with the number of frequency bins.  However this allows
for early termination of rays that are fully absorbed, which are
likely to have high absorption cross-sections (e.g. \ion{H}{1}
ionizations near 13.6 eV) or a low initial intensity (e.g. \ion{He}{2}
ionizing photons in typical stellar populations).  The other approach
used by some groups \citep[e.g.][]{Trac07} is to store all energy
groups in a single ray.  This reduces the number of the rays generated
and the computation associated with the ray tracing.  Unless the ray
dynamically adjusts its memory allocation for the energy groups as
they become depleted, this method is also memory intensive in the
situation where most of the energy groups are completely absorbed but
a few groups still have significant flux.

In practice, we have found that one energy group per photon type is
sufficient to match expected analytical tests (\S\ref{sec:nu_dep}).
For example when modeling Population III stellar radiation
\citep[e.g.][for hydrogen ionizing radiation only]{Abel07,
  Wise08_Gal}, we have 3 energy groups---\ion{H}{1}, \ion{He}{1},
\ion{He}{2}---each with an energy that equals the average photon
energy above the ionization threshold.

@@

\subsection{Test 14. \ion{H}{2} region with MHD}

>>>>

\documentclass[useAMS,usenatbib]{mn2e}

\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{natbib}
%\usepackage[section] {placeins}
\bibliographystyle{mn2e}
\citestyle{mn2e}

@@

\title[AMR Simulations with Adaptive Ray Tracing] {Enzo+Moray:
  Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with
  Adaptive Ray Tracing}

\author[J. H. Wise and T. Abel] {John H. Wise$^1$\thanks{Hubble
    Fellow; e-mail: jwise@astro.princeton.edu} and Tom Abel$^2$\\
$^{1}$ Department of Astrophysical Sciences, Princeton University,
Peyton Hall, Ivy Lane, Princeton, NJ 08544, USA\\
$^{2}$ Kavli Institute for Particle Astrophysics and Cosmology,
Stanford University, Menlo Park, CA 94025, USA}

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2010}

\maketitle
\label{firstpage}

@@

  We describe a photon-conserving numerical method that solves the
  radiative transfer equation, using a spatially-adaptive ray tracing
  scheme, and its parallel implementation into the adaptive mesh
  refinement (AMR) code, Enzo.  By coupling the solver with the energy
  equation and non-equilibrium chemistry network, our radiation
  hydrodynamics framework can be utilized to study a broad range of
  astrophysical problems, such as stellar and black hole (BH)
  feedback.  Inaccuracies can arise from large timesteps and poor
  sampling, therefore we devised an adaptive time-stepping scheme and
  a fast approximation of the optically-thin radiation field with
  multiple sources.  We test the method with several radiative
  transfer and radiation hydrodynamics tests that are given in
  \citet{RT06, Iliev09}.  We further test our method with more
  dynamical situations, for example, the propagation of an ionization
  front through a Rayleigh-Taylor instability, time-varying
  luminosities, and collimated radiation.  The test suite also
  includes an expanding \hii region in a magnetized medium,
  utilizing the newly implemented magnetohydrodynamics module in
  \enzo.  This method linearly scales with the number of point sources
  and number of grid cells.  To combat this, we use a novel method of
  merging rays at large radius, using a tree method, which we briefly
  describe.  Our implementation is scalable to 512 processors on
  distributed memory machines and can include radiation pressure and
  secondary ionizations from X-ray radiation.  It is included in the
  newest public release of Enzo.

@@

\begin{keywords}
  cosmology --- methods: numerical --- hydrodynamics --- radiative
  transfer.
\end{keywords}

@@

Radiative transfer is ubiquitous in many astrophysical problems,
encompassing topics such as stellar atmospheres, the interstellar
medium (ISM), star formation, galaxy formation, supernovae (SNe) and
cosmology.  Thus it is a well-studied problem
\citep[e.g.][]{Mathews65, Rybicki, Mihalas84, Yorke86}; however, its
treatment in multi-dimensional calculations is difficult because of
the dependence on seven variables -- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  In the general case, one must consider all radiation
sources at every point in calculation.

Fortunately problems can be simplified by exploiting symmetries and
special cases.  The first case was considered by \citet{Stroemgren39}
for a radiation source ionizing a static uniform neutral medium, where
recombinations balance photo-ionizations.  He derived a radius of an
\hii region,
%
\begin{equation}
  \label{eqn:Rstr}
  R_s = \left(\frac{3\dot{N}_\gamma}{4\pi \alpha_{\rm B} n_H^2}\right)^{1/3},
\end{equation}
where $\dot{N}_\gamma$ is the ionizing photon luminosity, $\alpha_{\rm
  B}$ is the recombination rate, and $n_H$ is the ambient hydrogen
number density.  Furthermore he found that the delineation between the
neutral and ionized medium to be approximately the mean free path of
the ionizing radiation.  His seminal work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionizing radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionization front,
then this over-pressurized region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics.  A
decade later, the first radiation hydrodynamical numerical models of
\hii regions in spherical symmetry and plane-parallel ionization
fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionization front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  In R-type
fronts, the ionized gas density is higher than the neutral gas
density, and in D-type fronts, the opposite is true.  R-type fronts
travel supersonically with respect to the neutral gas, whereas D-type
fronts are subsonic.  Furthermore ``weak'' and ``strong'' R-type fronts
move supersonically and subsonically with respect to the ionized gas,
respectively.  The same terminology conversely applies to D-type
fronts.  ``Critical'' fronts are defined as moving exactly at the
sound speed.  These works established the evolutionary track of an
expanding \hii illuminated by a massive star in a uniform
medium:
%
\begin{enumerate}
\item Weak R-type:  When the star (gradually) starts to shine, the
  ionization front will move supersonically through the ambient
  medium.  The gas is heated and ionized, but otherwise left
  undisturbed.  This stage continues until $r \sim 0.02R_s$.
\item Critical R-type:  As the ionization front moves outwards, it
  begins to slow because of the geometric dilution of the radiation.
  It becomes a critical R-type front, which is equivalent to an
  isothermal shock in the neutral gas.
\item Strong and weak D-type:  The front continues to slow, becoming a
  strong D-type front, and then a critical D-type front.  From this
  point forward, it is moving subsonically with respect to the ionized
  gas, i.e. a weak D-type front.  Thus sound waves can travel across
  the ionization front and form a shock.  The ionization front
  detaches from the shock, putting the shock ahead of the ionization
  front.
\item Expansion phase:  After the shock forms, the \hii region
  starts to expand, lowering the interior density and thus the
  recombination rates.  This increases the number of photons available
  for ionizing the gas.  The sphere expands until it reaches pressure
  equilibrium with the ambient medium at $r \sim 5R_s$.
\end{enumerate}
%

In the 1970's and 1980's, algorithmic and computational advances
allowed numerical models to be expanded to two dimensions, mainly
using axi-symmetric to simplify the problem
\citep[e.g.][]{Bodenheimer79, Sandford82, Yorke83}.  One topic that was
studied extensively were champagne flows.  Here the source is embedded
in an overdense region, and the \hii region escapes from this
region in one direction.  The interface between the ambient and dense
medium was usually set up to be a constant pressure boundary.  When
the ionization front passes this boundary, the dense, ionized gas is
orders of magnitude out of pressure equilibrium as the temperatures on
both sides of initial boundary are within a factor of a few.  In
response, the gas is accelerated outwards in this direction, creating
a fan-shaped outflow.

Only in the past 15 years, computational resources have become large
enough, along with further algorithmic advances, to cope with the
requirements of three-dimensional calculations.  There are two popular
methods to solve the radiative transfer equation in three-dimensions:
%
\begin{itemize}
\item Moment methods:  The angular moments of the radiation field can
  describe its angular structure, which are related to the energy
  energy, flux, and radiation pressure \citep{Auer70, Norman98}.
  These have been implemented in conjunction with short
  characteristics \citep[2D]{Stone92_RHD}, with long characteristics
  \citep{Finlator09}, with a variable Eddington tensor in the
  optically thin limit \citep[OTVET;][]{Gnedin01_OTVET, Petkova09},
  and with an M1 closure relation \citep{Gonzalez07, Aubert08}.
  Moment methods have the advantage of being fast, being independent
  of the number of radiation point sources.  However, they are
  diffusive and result in incorrect shadows in some situations.
\item Ray tracing:  Radiation can be propagated along rays that extend
  through the computational grid \citep[e.g.][]{Razoumov99, Abel99_RT,
    Ciardi01, Sokasian01, Whalen06, Rijkhorst06, Mellema06, Alvarez06,
    Trac07, Krumholz07_ART, Paardekooper10} or particle set
  \citep[e.g.][]{Susa06, Johnson07, Pawlik08, Pawlik10, Altay08,
    Hasegawa09}.  In general, these methods are very accurate but
  computationally expensive because the radiation field must be well
  sampled by the rays with respect to the spatial resolution of the
  grid or particles.
\end{itemize}
%
Until the mid-2000's the vast majority of the three-dimensional
calculations were performed with static density distributions.  One
example is calculating cosmological reionization by post-processing of
density fields from N-body simulations \citep{Ciardi01, Sokasian01,
  McQuinn07, Iliev06, Iliev07}.  Any hydrodynamical response to the
radiation field was thus ignored.  Several radiative transfer codes
were compared in four purely radiative transfer tests in
\citet[hereafter RT06]{RT06}.  Only recently has the radiative
transfer equation been coupled to the hydrodynamics in
three-dimensions.  In the second comparison paper \citep[hereafter
  RT09]{Iliev09}, results from these radiation hydrodynamics codes
were compared.  Even more rare are ones that couple it with
magneto-hydrodynamics \citep[e.g.][]{Krumholz07_ART}.  The tests in
RT06 and RT09 were kept relatively simple to ease the comparison.

In this paper, we present our implementation, Enzo+Moray, of adaptive
ray tracing \citep{Abel02_RT} in the cosmological hydrodynamics
adaptive mesh refinement (AMR) code, \enzo~\citep{BryanNorman1997,
  OShea2004}.  The radiation field is coupled to the hydrodynamics
solver at small timescales, enabling it to study radiation
hydrodynamical problems.  We have used this code to investigate the
growth of an \hii region from a 100\Ms~Population III (Pop III)
star \citep{Abel07}, the early stages of reionization from Pop III
stars \citep{Wise08_Reion}, radiative feedback on the formation of
high redshift dwarf galaxies \citep{Wise08_Gal}, ultraviolet radiation
escape fractions from dwarf galaxies before reionization
\citep{Wise09}, negative radiative feedback from accreting Pop III
seed black holes \citep{Alvarez09}, and radiative feedback in
accreting supermassive black holes \citep[][in prep.]{Kim11}.  We have
included Enzo+Moray in the latest public release of
\enzo\footnote{http://lca.ucsd.edu/enzo}, and it is also coupled with
the newly added MHD solver in \enzo~\citep{Wang09}.

@@

\input table1_mnras.tex

@@

\begin{figure}
  \plottwo{fig/covering.eps}{fig/calc_fc.eps}
  \caption{\label{fig:covering} (a) A two-dimensional illustration of
    the overlap between the beam associated with a ray $\gamma$ and a
    computational cell.  The ray has a segment length of $dr$ passing
    through the cell.  The covering area is denoted by dark grey,
    where the full area $(dr \times L_{\rm pix})$ is colored by the
    dark and light grey.  The photo-ionization and photo-heating rates
    should be corrected by this overlap fraction $f_c$.  (b)
    Annotation of quantities used in this geometric correction.}
\end{figure}

@@

We must be careful not the overestimate the radiation when multiple
rays enter a single cell.  In the case of a single radiation source,
the solution is simple -- only assign the cell the photo-ionization
and photo-heating rates when $k_{\rm ph} = 0$.  However in the case
with multiple sources, this is no longer valid, and we must sum the
flux from all optically thin sources.  Only one ray per source must
contribute to a single cell in this framework.  We create a flagging
field that marks whether a cell has already been touched by an
optically thin photon from a particular radiation source.  Naively, we
would be restricted to tracing rays from a single source at a time if
we use a boolean flagging field.  However we can trace rays for 32
sources at a time by using bitwise operations on a 32-bit integer
field.  For example in \texttt{C}, we would check if an optically thin
ray from the \texttt{n}-th source has propagated through cell
\texttt{i} by evaluating \texttt{(MarkerField[i] $\gg$ n \& 1)}.  If
false, then we can add the optically thin approximation
(Eq. \ref{eqn:optthin}) to the cell and set \texttt{MarkerField[i] |=
  (1 $\gg$ n);} to mark the cell.

@@

Molecular hydrogen can absorb photons in the Lyman-Werner bands
through the two-step Solomon process that are composed of 76
absorption lines ranging from 11.1 to 13.6 eV \citep{Stecher67,
  Dalgarno70}.  Each of these spectral lines can be modelled with a
typical exponential attenuation equation \citep{Ricotti01}, but
\citet{Draine96} showed that this self-shielding is well modeled with
the following relation to total \hh~column density
%
\begin{equation}
  \label{eqn:lwband}
  f_{\rm shield} (N_{\rm H2}) =
  \left\{ \begin{array}{l@{\quad}l}
      1 & (N_{\rm H2} \le 10^{14})\\
      (N_{\rm H2}/10^{14})^{-0.75} & (N_{\rm H2} >
      10^{14})
    \end{array} \right. ,
\end{equation}
where $N_{\rm H2}$ is in units of cm$^{-2}$.  To incorporate this
shielding function into the ray tracer, we store the total \hh~column
density and calculate the \hh~dissociation rate by summing the
contribution of all rays,
%
\begin{equation}
  \label{eqn:lwRT}
  k_{\rm diss} = \sum_{\rm rays} \frac{P \; \sigma_{\rm LW} \;
    \Omega_{\rm ray} \; r^2 \; dr}{A_{\rm cell} \; dV \; dt_P},
\end{equation}
where $\sigma_{\rm LW} = 3.71 \times 10^{18} \mathrm{cm}^2$ is the
effective cross-section of \hh~\citep{Abel97}.  To account for
absorption, we attenuate the photon number flux by
%
\begin{equation}
  \label{eqn:LWdP}
  dP = P [f_{\rm shield}(N_{\rm H2} + dN_{\rm H2}) - f_{\rm shield}(N_{\rm H2})],
\end{equation}
where $dN_{\rm H2}$ is the \hh~column density in the current cell.

@@

At the other end of the spectrum, a high-energy ($E_{\rm ph} \gsim
100$ eV) photon can ionize multiple neutral hydrogen and helium atoms,
and this should be considered in such radiation fields.
\citet{Shull85} studied this effect with Monte Carlo calculations over
varying electron fractions and photon energies up to 3 keV.  They find
that the excitation of hydrogen and helium and the ionization of
He {\sc ii} is negligible.  The number of secondary ionizations of H
and He is reduced from the ratio of the photon and ionization energies
($E_{\rm ph} / E_i$) by a factor of
%
\begin{equation}
  Y_{\rm k,H} = 0.3908 (1 - x^{0.4092})^{1.7592},
\end{equation}
\begin{equation}
  Y_{\rm k,He} = 0.0554 (1 - x^{0.4614})^{1.6660},
\end{equation}
where $x$ is the electron fraction.  The remainder of the photon
energy is deposited into thermal energy that is approximated by
%
\begin{equation}
  Y_{\rm \Gamma} = 0.9971 [ 1 - (1 - x^{0.2663})^{1.3163} ]
\end{equation}
and approaches one as $x \rightarrow 1$.  Thus in gas with low
electron fractions, most of the energy results in ionizations of
hydrogen and helium, and in nearly ionized gas, the energy goes into
photo-heating.

@@

\begin{figure*}
  \epsscale{2}
  \plotone{fig/RTcoupling.eps}
  \epsscale{1}
  \caption{\label{fig:MainAlgorithm} Flow chart for the overall
    algorithm of the radiative transfer module in \enzo~that
    illustrates (1) the creation of photon packages, (2) ray tracing,
    (3) the transport of photon packages between AMR grids, and (4)
    coupling with the hydrodynamics.  The ray tracing algorithm, which
    is contained in the ``Trace Rays'' is detailed in Figure
    \ref{fig:RTalgorithm}.}
\end{figure*}

\begin{figure*}
  \epsscale{2}
  \plotone{fig/RTalgorithm.eps}
  \epsscale{1}
  \caption{\label{fig:RTalgorithm} Flow chart for the ray tracing
    algorithm for one photon passing through a grid.  Note that only
    one step is needed in the routine to adaptively split rays.  The
    remainder is a typical ray tracing method.}
\end{figure*}

@@

\step{3g} Add the effects of any optional physics modules
(\S\ref{sec:addphysics}) -- secondary ionizations from X-rays, Compton
heating from scattering, and radiation pressure.

@@

In our implementation, photon packages are mono-chromatic, i.e. energy
groups \citep[][Ch. 6]{Mihalas84}, and are assigned a photon type that
corresponds whether it is a photon that (1) ionizes hydrogen, (2)
singly ionizes helium, (3) doubly ionizes helium, (4) has an X-ray
energy, or (5) dissociates molecular hydrogen (Lyman-Werner
radiation).  One disadvantage of mono-chromatic rays is the number of
rays increase with the number of frequency bins.  However this allows
for early termination of rays that are fully absorbed, which are
likely to have high absorption cross-sections (e.g. \hi ionizations
near 13.6 eV) or a low initial intensity (e.g. He {\sc ii} ionizing
photons in typical stellar populations).  The other approach used by
some groups \citep[e.g.][]{Trac07} is to store all energy groups in a
single ray.  This reduces the number of the rays generated and the
computation associated with the ray tracing.  Unless the ray
dynamically adjusts its memory allocation for the energy groups as
they become depleted, this method is also memory intensive in the
situation where most of the energy groups are completely absorbed but
a few groups still have significant flux.

In practice, we have found that one energy group per photon type is
sufficient to match expected analytical tests (\S\ref{sec:nu_dep}).
For example when modeling Population III stellar radiation
\citep[e.g.][for hydrogen ionizing radiation only]{Abel07,
  Wise08_Gal}, we have 3 energy groups -- H {\sc i}, He {\sc i}, He
{\sc ii} -- each with an energy that equals the average photon energy
above the ionization threshold.

@@

\subsection{Test 14. \hii region with MHD}

\begin{figure*}
  \epsscale{2}
  \plotone{fig/test14_slices_xy}
  \epsscale{1}
  \caption{\label{fig:test14_1} Test 14. (\hii region with MHD).
    Left to right: slices of density at $t = 0.18, 0.53, 1.58$ Myr in
    the x-y plane.  The streamlines show the magnetic field.}
\end{figure*}
